Multiparty Session Type-safe Web Development in TypeScript

### 2020-01-21 PLACES_2020/paper.tex
PLACES 2020: check TexMaker pdflatex functionality
PLACES 2020: complete first draft

\begin{abstract}
% Situation: webapps, interactive, JavaScript more powerful but remains a poor language for maintaing large applications

% Problem: hard to maintain/test
% New languages (like Dart and TypeScript)

% Solution: MPST for web development

This is a sentence in the abstract.
This is another sentence in the abstract.
This is yet another sentence in the abstract.
This is the final sentence in the abstract.

% Problem: similar to PureScript paper

% In this paper, we present...

% TypeScript
% - gradual type system, superset of JavaScript
% - we define how our approach achieves "gradual linearity" (can we define this?)
%Our work targets TypeScript - a superset of \dots

%We demonstrate the effectiveness of our approach through a web-based game of \textit{Noughts and Crosses} where communication is performed through WebSocket connections.

@@

% Example: naughts and crosses
Consider a classic turn-based board game of \textit{Noughts and Crosses} between two players. Both players are identified by either noughts or crosses respectively, and take turns to place a mark on an unoccupied cell of a 3-by-3 grid until one player wins (when their markers form one straight line on the board) or a stalemate is reached (when all cells are occupied). A web-based implementation may involve players connected to a game server via WebSocket connections and interacting with the game from their web browser, which serve a \textit{single-page application} (SPA) of the game client written in a popular framework like \textit{React.js} \cite{React}. SPAs feature a single HTML page and dynamically renders content via JavaScript in the browser. Players take turns to make a move on the game board and the server implements the game logic to progress the game forward until a result (either a win/loss or draw) can be declared.

@@

\subsubsection{Session Runtime and Protocol API}
\label{section:serversessionapi}

Our session runtime performs communication in a way that conforms to the protocol specification. The runtime listens to message (receive) events on the communication channel, invokes the corresponding callback to obtain the value to send next, and performs the send.

The \textit{Protocol API} is a class that interfaces between the session runtime and the user implementation of the EFSM state encodings. Upon construction, the Protocol class waits for all client roles to connect via WebSocket, then proceeds to the initial state of the EFSM and performs state transitions depending on the permitted communication patterns from the EFSM.

@@

\paragraph{Unused} The initial state must be supplied to the Protocol API constructor in order to instantiate the session; this initial state is defined in terms of the successor states, which in turn has references to its successors and so forth. This encoding approach will cover the terminal state (if it exists in the EFSM, as \cite{Hybrid2016} notes that an EFSM contains at most one terminal state), and the session runtime guarantees this terminal state, if it exists, will be reached by construction.

>>>>

\begin{abstract}
Advancements in mobile device compute power has made interactive web applications possible, allowing the web browser to render content dynamically and support low-latency communication with the server. This comes at a cost to the developer, who now need to reason more about the communication correctness of their application as they support more complex communication patterns.

Multiparty session types (MPST) provide a framework for verifying implementations to conform to a prescribed communication protocol. Existing proposals for applying the MPST framework in application development either neglect the event-driven nature of web applications, or lack compatibility with industry tools and practices which discourage mainstream adoption by web developers.

In this paper, we present an implementation of the MPST framework for developing interactive web applications using familiar industry tools in TypeScript and the \textit{React.js} framework. The global communication protocol is specified in the Scribble protocol language and checked for well-formedness by the Scribble toolchain. We obtain the local protocol for each role, which describes the interactions of the global communication protocols observed by the role. We encode the local protocol into TypeScript types, catering for server-side and client-side targets separately, and show that our encoding guarantees that only implementations which conform to the protocol will type-check. We demonstrate the effectiveness of our approach through a web-based implementation of the classic \textit{Noughts and Crosses} game from a MPST formalism of the game logic.

@@

% Example: noughts and crosses
Consider a classic turn-based board game of \textit{Noughts and Crosses} between two players. Both players are identified by either noughts or crosses respectively, and take turns to place a mark on an unoccupied cell of a 3-by-3 grid until one player wins (when their markers form one straight line on the board) or a stalemate is reached (when all cells are occupied). A web-based implementation may involve players connected to a game server via WebSocket connections and interacting with the game from their web browser, which serve a \textit{single-page application} (SPA) of the game client written in a popular framework like \textit{React.js} \cite{React}. SPAs feature a single HTML page and dynamically renders content via JavaScript in the browser. Players take turns to make a move on the game board and the server implements the game logic to progress the game forward until a result (either a win/loss or draw) can be declared.

@@

\subsubsection{Session runtime}
\label{section:serversessionapi}

Our session runtime performs communication in a way that conforms to the protocol specification, but does not expose these IO actions to the developer by delegating the aforementioned responsibilities to an inner class. The runtime listens to message (receive) events on the communication channel, invokes the corresponding callback to obtain the value to send next, and performs the send. The developer instantiates the session by constructing an instance of the session runtime class.

@@

\paragraph{Unused} The initial state must be supplied to the session runtime constructor in order to instantiate the session; this initial state is defined in terms of the successor states, which in turn has references to its successors and so forth. This encoding approach will cover the terminal state (if it exists in the EFSM, as \cite{Hybrid2016} notes that an EFSM contains at most one terminal state), and the session runtime guarantees this terminal state, if it exists, will be reached by construction.
